The Calculus of Linear Constructions

### 2022-01-08 writing/appendix/draft.tex
draft

\begin{align*}
  occurs\ x\ y                               & =
  \begin{dcases}
    1 & x =_\alpha y    \\
    0 & x \neq_\alpha y
  \end{dcases}                                                \\
  occurs\ x\ s_i                             & = 0                           \\
  occurs\ x\ ((y \stype{s} A) \rightarrow B) & = occurs\ x\ A + occurs\ x\ B \\
  occurs\ x\ ((y \stype{s} A) \multimap B)   & = occurs\ x\ A + occurs\ x\ B \\
  occurs\ x\ (\lambda x \stype{s} A.n)       & = occurs\ x\ A + occurs\ x\ n \\
  occurs\ x\ (m\ n)                          & = occurs\ x\ m + occurs\ x\ n
\end{align*}

@@

\begin{lemma}\label{narity}
  For context $\Gamma$, terms $m, A$, if there is $\Gamma \vdash m : A$, then for any variable $x \notin \Gamma$ there is $occurs\ x\ m = 0$
\end{lemma}
\begin{proof}
  By induction on the derivation of $\Gamma \vdash m : A$.
\end{proof}

\begin{theorem}\label{linearity}
  Linearity. For context $\Gamma$, terms $m, A$, if there is $\Gamma \vdash m : A$, then for any variable with linear type $x \in \Gamma$ there is $occurs\ x\ m = 1$.
\end{theorem}
\begin{proof}
  The proof proceeds by induction on the derivation of $\Gamma \vdash m : A$, we will discuss the application cases in detail.
  \begin{itemize}
    \item For case \rname{App-U$\rightarrow$}, by assumptions $\mrg{\Gamma_1}{\Gamma_2}{\Gamma}$ and $(x \ltype A) \in \Gamma$ and Lemma \ref{islmergeinv} we can conclude that $x \in \Gamma_1$ and $x \notin \Gamma_2$ or $x \in \Gamma_2$ and $x \notin \Gamma_1$. In both cases, applying the induction hypothesis and Lemma \ref{narity} proves the goal.
    \item For case \rname{App-L$\rightarrow$}, the proof is the same as \rname{App-U$\rightarrow$}.
    \item For case \rname{App-U$\multimap$}, the proof is the same as \rname{App-U$\rightarrow$}.
    \item For case \rname{App-L$\multimap$}, the proof is the same as \rname{App-U$\rightarrow$}.
  \end{itemize}
\end{proof}

>>>>

\begin{align*}
  occurs(x, y)                               & =
  \begin{dcases}
    1 & x =_\alpha y    \\
    0 & x \neq_\alpha y
  \end{dcases}                                                \\
  occurs(x, s_i)                             & = 0                           \\
  occurs(x, ((y \stype{s} A) \rightarrow B)) & = occurs(x, A) + occurs(x, B) \\
  occurs(x, ((y \stype{s} A) \multimap B))   & = occurs(x, A) + occurs(x, B) \\
  occurs(x, (\lambda x \stype{s} A.n))       & = occurs(x, A) + occurs(x, n) \\
  occurs(x, (m\ n))                          & = occurs(x, m) + occurs(x, n)
\end{align*}

@@

\begin{lemma}\label{narity}
  For context $\Gamma$, terms $m, A$, if there is $\Gamma \vdash m : A$, then for any variable $x \notin \Gamma$ there is $occurs(x, m) = 0$
\end{lemma}
\begin{proof}
  By induction on the derivation of $\Gamma \vdash m : A$.
\end{proof}

\begin{theorem}\label{linearity}
  Linearity. For context $\Gamma$, terms $m, A$, if there is $\Gamma \vdash m : A$, then for any variable with linear type $x \in \Gamma$ there is $occurs(x, m) = 1$.
\end{theorem}
\begin{proof}
  The proof proceeds by induction on the derivation of $\Gamma \vdash m : A$, we will discuss the application cases in detail.
  \begin{itemize}
    \item For case \rname{App-U$\rightarrow$}, by assumptions $\mrg{\Gamma_1}{\Gamma_2}{\Gamma}$ and $(x \ltype A) \in \Gamma$ and Lemma \ref{islmergeinv} we can conclude that $x \in \Gamma_1$ and $x \notin \Gamma_2$ or $x \in \Gamma_2$ and $x \notin \Gamma_1$. In both cases, applying the induction hypothesis and Lemma \ref{narity} proves the goal.
    \item For case \rname{App-L$\rightarrow$}, the proof is the same as \rname{App-U$\rightarrow$}.
    \item For case \rname{App-U$\multimap$}, the proof is the same as \rname{App-U$\rightarrow$}.
    \item For case \rname{App-L$\multimap$}, the proof is the same as \rname{App-U$\rightarrow$}.
  \end{itemize}
\end{proof}

### 2022-01-08 writing/draft6/draft6.tex
draft

However, due to its origins as a logical framework for constructive mathematics, it is quite difficult for CC to encode and reason about resources. Users of proof assistants based on CC often need to embed external logics such as Separation Logic to provide domain-specific reasoning principles for dealing with resource. We propose an alternative solution: extend CC with linear types.

@@

We further extend CLC with rules for defining inductive linear types in the style of CIC, forming the Calculus of Inductive Linear Constructions (CILC). Linear connectives such a $\otimes$ and $\oplus$ are definable using this generalized mechanism in a straightforward way. Through examples, we show that CILC facilitates lightweight reasoning about stateful programs and protocol enforcement.

@@

\noindent \textbf{\textit{Contributions}}:
Our contributions can be summarized as follows.
\begin{itemize}
  \item Fist, we describe the Calculus of Linear Constructions, an extension to the Calculus of Constructions with linear types. The integration of linear types and dependent types allows CLC to directly and precisely reason about resource.
  \item Second, we study the meta-theory of CLC directly, showing that it satisfies the standard properties of confluence, validity and subject reduction.
  \item Next, we prove the linearity theorem, showing the tangible impact linear types have on the structure of terms. We also show that promotion and dereliction of linear logic are derivable as theorems for canonical types.
  \item Additionally, we observe that CLC is compatible with the predicative CC$\omega$, allowing us to construct a reduction preserving model of CLC in CC$\omega$, showing that CLC is consistent. We prove that all valid CC$\omega$ terms are valid CLC terms after adding minimal amounts of annotation.
  \item Furthermore, we extend CLC with rules to define inductive linear types in the style of CIC, forming the Calculus of Inductive Linear Constructions (CILC). This is the first presentation of a generalized mechanism for defining inductive linear types that we are aware of.
  \item All major results have been formalized and proven correct in the Coq Proof Assistant with help from the Autosubst \cite{autosubst} library. To the best of our knowledge, our development is the first machine checked formalization of a linear dependent type theory.
  \item Finally, we give an implementation of CILC in OCaml. Algorithmic type checking employed by the typechecker streamlines the process of writing CILC programs.
\end{itemize}

@@

In \rname{$\lambda$$\rightarrow$}, the function type being addressed has form $(x \stype{s} A) \rightarrow B$. $\lambda$-abstractions of this type can be applied an unrestricted number of times, hence cannot depend on free variables with restricted usage without possibly duplicating them. This consideration is realized by the side condition $\pure{\Gamma}$, asserting all variables in context $\Gamma$ are unrestricted. Next, the body of the abstraction $n$ is typed as $\Gamma, x \stype{s} A \vdash n : B$, where $s$ is the sort of $A$. Finally, the resulting judgment $\Gamma \vdash \lambda x.n : (x \stype{s} A) \rightarrow B$ asserts that the $\lambda$-abstraction can be used unrestrictedly.

In contrast to \rname{$\lambda$$\rightarrow$}, the \rname{$\lambda$$\multimap$} rule is used for forming $\lambda$-abstractions that must be used exactly once. Due to the fact these abstractions must be used once, they are allowed access to restricted variables within context $\Gamma$, evident in the judgment $\Gamma, x \stype{s} A \vdash n : B$, and lack of side condition $\pure{\Gamma}$. However, in judgment $\overline{\Gamma} \vdash (x \stype{s} A) \multimap B : L_k$ the context must be filtered, because types are not allowed to depend on restricted variables. The final resulting judgment $\Gamma \vdash \lambda x.n : (x \stype{s} A) \multimap B$ asserts that the $\lambda$-abstraction must be used exactly once.

@@

All proofs have been formalized in Coq with help from the Autosubst \cite{autosubst} library. The Coq development is publicly available on the first author's Github repository. To the best of our knowledge, this is the first machined checked formalization of a linear dependently type theory. We give a hand written version of the proof in the appendix as well.

@@

\subsubsection{Substitution} \label{subst}
Though the substitution lemma is regarded as a boring and bureaucratic result, it is surprisingly hard to design linear typed languages where the substitution lemma is admissible. Much of the difficulty arises during the substitution of arguments containing ! exponential. Perhaps the most famous work detailing the issues of substitution is due to Wadler \cite{substitute}. He defines additional syntax and semantics for the intricate unboxing of ! terms, solving the lack of substitute in Abramsky's term calculus.

>>>>

However, due to its origins as a logical framework for constructive mathematics, it can be inconvenient for CC to encode and reason about resources. Users of proof assistants based on CC often need to embed external logics such as Separation Logic to provide domain-specific reasoning principles for dealing with resource. We propose an alternative solution: extend CC with linear types.

@@

We further extend CLC with rules for defining inductive linear types in the style of CIC, forming the Calculus of Inductive Linear Constructions (CILC). Standard Linear Logic connectives such a $\otimes$ and $\oplus$ are definable using this generalized mechanism in a straightforward way. Through examples, we show that CILC facilitates lightweight reasoning about stateful programs and protocol enforcement.

@@

\noindent \textbf{\textit{Contributions}}:
Our contributions can be summarized as follows.
\begin{itemize}
  \item Fist, we describe the Calculus of Linear Constructions, an extension to the Calculus of Constructions with linear types. The integration of linear types and dependent types allows CLC to directly and precisely reason about resource.
  \item Second, we study the meta-theory of CLC directly, showing that it satisfies the standard properties of confluence, validity and subject reduction.
  \item Next, we prove the linearity theorem, showing the tangible impact linear types have on the structure of terms. We also show that promotion and dereliction of linear logic are derivable as theorems for function types.
  \item Additionally, we observe that CLC is compatible with the predicative CC$\omega$, allowing us to construct a reduction preserving model of CLC in CC$\omega$, showing that CLC is consistent. We prove that all valid CC$\omega$ terms are valid CLC terms after adding minimal amounts of annotation.
  \item Furthermore, we extend CLC with rules to define inductive linear types in the style of CIC, forming the Calculus of Inductive Linear Constructions (CILC). This is the first presentation of a generalized mechanism for defining inductive linear types that we are aware of.
  \item All major results have been formalized and proven correct in the Coq Proof Assistant with help from the Autosubst \cite{autosubst} library. To the best of our knowledge, our development is the first machine checked formalization of a linear dependent type theory.
  \item Finally, we give an implementation of CILC in OCaml. Algorithmic type checking employed by the typechecker streamlines the process of writing CILC programs.
\end{itemize}

@@

In \rname{$\lambda$$\rightarrow$}, the function type has form $(x \stype{s} A) \rightarrow B$. Abstractions of this type can be applied an unrestricted number of times, hence cannot depend on free variables with restricted usage without possibly duplicating them. This consideration is realized by the side condition $\pure{\Gamma}$, asserting all variables in context $\Gamma$ are unrestricted. Next, the body of the abstraction $n$ is typed as $\Gamma, x \stype{s} A \vdash n : B$, where $s$ is the sort of $A$. Finally, the resulting judgment $\Gamma \vdash \lambda x.n : (x \stype{s} A) \rightarrow B$ asserts that the $\lambda$-abstraction can be used unrestrictedly.

In contrast to \rname{$\lambda$$\rightarrow$}, the \rname{$\lambda$$\multimap$} rule is used for forming abstractions that must be used exactly once. Due to the fact these abstractions must be used once, they are allowed access to restricted variables within context $\Gamma$, evident in the judgment $\Gamma, x \stype{s} A \vdash n : B$, and lack of side condition $\pure{\Gamma}$. However, in judgment $\overline{\Gamma} \vdash (x \stype{s} A) \multimap B : L_k$ the context must be filtered, because types are not allowed to depend on restricted variables. The final resulting judgment $\Gamma \vdash \lambda x.n : (x \stype{s} A) \multimap B$ asserts that the $\lambda$-abstraction must be used exactly once.

@@

All proofs have been formalized in Coq with help from the Autosubst \cite{autosubst} library. The Coq development is publicly available on the first author's Github repository. To the best of our knowledge, this is the first machined checked formalization of a linear dependently type theory. We give a hand written version of the proof in an accompanying technical report.

@@

\subsubsection{Substitution} \label{subst}
Though the substitution lemma is regarded as a boring and bureaucratic result, it is surprisingly hard to design linear typed languages where the substitution lemma is admissible. Much of the difficulty arises during the substitution of arguments containing ! exponential. Perhaps the most famous work detailing the issues of substitution is due to Wadler \cite{substitute}. He defines intricate syntax and semantics for the unboxing of ! terms, solving the lack of substitute in Abramsky's term calculus.
